Stochastic Variance Reduced Policy Gradient

### 2018-02-04 svrpg.tex
added appendix on applicability to Gaussian policies

\section{Convergence Guarantees}\label{sec:conv}
In this section we adapt existing convergence guarantees for nonconvex SVRG \cite{reddi2016stochastic} \cite{allen2016variance} to the Policy Gradient framework.
Each of the three challenges presented at the beginning of Section \ref{sec:alg} can potentially prevent convergence, so we need additional assumptions:
\begin{enumerate}
\item \textit{Non-convexity}:
	as already mentioned, a common assumption for nonconvex optimization is the L-smoothness of the objective. We actually require the following, which is stronger:
	\begin{assumption}\label{ass:bounded_score}
		For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
		\begin{align*}
		&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
		&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
		\end{align*}
	\end{assumption}
	This property is satisfied by the commonly used truncated Gaussian policy, and is a sufficient condition for the L-smoothness of the objective (see Lemma \ref{lemma:lsmooth} in the appendix).
\item \textit{Approximation}:
since we can only compute an estimate of the full gradient, we require its variance to be bounded:
	\begin{assumption}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for each policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
	\end{assumption}
This assumption is satisfied by the REINFORCE and G(PO)MDP estimators and Gaussian policies \cite{zhao2011analysis} \cite{pirotta2013adaptive}. The problem of full gradient approximation for SVRG has been addressed in the supervised learning framework \cite{harikandeh2015stopwasting} with results similar to ours.
\item \textit{Non-stationarity}:
the variance introduced by importance sampling must be bounded:
	\begin{assumption}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
		\[
		\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS.
		\]
	\end{assumption}
Differently from the previous two, this assumption must be enforced by a proper handling of the epoch size $m$.
\end{enumerate}

We can now state the following key result:
\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score}, \ref{ass:REINFORCE} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size $\alpha$ and the epoch size $m$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B}
\end{align*}
\end{restatable}
Specific constants and meta-parameter constraints are provided in the proof of the theorem in Appendix \ref{app:proofs}. In Section \ref{sec:stopping}, instead, we propose a joint selection of step size $\alpha$ and epoch size $m$ which is more suited for applications. 
The $\mathcal{O}(\nicefrac{1}{T})$ term is coherent with results on non-convex SVRG, \eg \cite{reddi2016stochastic}; the $\mathcal{O}(\nicefrac{1}{N})$ term is due to full gradient approximation and is analogous to the one in \cite{harikandeh2015stopwasting}; the $\mathcal{O}(\nicefrac{1}{B})$ term is due to importance weighting. To achieve true convergence, both the batch size $N$ and the mini-batch size $B$ should increase over time. In practice, it is enough to choose $N$ and $B$ large enough to make the second and the third term negligible, \ie to mitigate the variance introduced by full gradient approximation and importance sampling respectively. Since the optimal configuration may be task-dependent, we leave this choice to the empirical analysis of Section \ref{sec:exp}.

@@

The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\left(\sum_{h=0}^{H-1}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H-1}\gamma^h r_h^n - b\right),
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline, provided it is constant \wrt to actions.
\todopir{Actually it can be state-action dependant~\citep[\eg][]{Thomas2017actionbaseline,wu2018variance}.}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r_h^n - b\right).
\end{align*}

@@

\section{Applicability to Gaussian Policies}\label{app:pol}
\todomat{...}

>>>>

\section{Convergence Guarantees}\label{sec:conv}
In this section we adapt existing convergence guarantees for nonconvex SVRG \cite{reddi2016stochastic} \cite{allen2016variance} to the Policy Gradient framework.
Each of the three challenges presented at the beginning of Section \ref{sec:alg} can potentially prevent convergence, so we need additional assumptions:
\begin{enumerate}
\item \textit{Non-convexity}:
	as already mentioned, a common assumption for nonconvex optimization is the L-smoothness of the objective. We actually require the following, which is stronger:
	\begin{restatable}[On policy derivatives]{assumption}{boundedscore}\label{ass:bounded_score}
		For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
		\begin{align*}
		&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
		&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
		\end{align*}
	\end{restatable}
	This property is a sufficient condition for the L-smoothness of the objective (see Lemma \ref{lemma:lsmooth} in the appendix).
\item \textit{Approximation}:
since we can only compute an estimate of the full gradient, we require its variance to be bounded:
	\begin{restatable}[On the variance of the gradient estimator]{assumption}{varreinforce}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for each policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
	\end{restatable}
The problem of full gradient approximation for SVRG has been addressed in the supervised learning framework \cite{harikandeh2015stopwasting} with results similar to ours.
\item \textit{Non-stationarity}:
the variance introduced by importance sampling must be bounded:
	\begin{restatable}[On the variance of importance weights]{assumption}{varweights}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
		\[
		\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS.
		\]
	\end{restatable}
Differently from the previous two, this assumption can be enforced by a proper handling of the epoch size $m$.
\end{enumerate}

In Appendix \ref{app:gauss} we show how the commonly used Gaussian policy satisfies these assumptions.
We can now state the following key result:
\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score}, \ref{ass:REINFORCE} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size $\alpha$ and the epoch size $m$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B}
\end{align*}
\end{restatable}
Specific constants and meta-parameter constraints are provided in the proof of the theorem in Appendix \ref{app:proofs}. In Section \ref{sec:stopping}, instead, we propose a joint selection of step size $\alpha$ and epoch size $m$ which is more suited for applications. 
The $\mathcal{O}(\nicefrac{1}{T})$ term is coherent with results on non-convex SVRG, \eg \cite{reddi2016stochastic}; the $\mathcal{O}(\nicefrac{1}{N})$ term is due to full gradient approximation and is analogous to the one in \cite{harikandeh2015stopwasting}; the $\mathcal{O}(\nicefrac{1}{B})$ term is due to importance weighting. To achieve true convergence, both the batch size $N$ and the mini-batch size $B$ should increase over time. In practice, it is enough to choose $N$ and $B$ large enough to make the second and the third term negligible, \ie to mitigate the variance introduced by full gradient approximation and importance sampling respectively. Since the optimal configuration may be task-dependent, we leave this choice to the empirical analysis of Section \ref{sec:exp}.

@@

The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\left(\sum_{h=0}^{H-1}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H-1}\gamma^h r_h^n - b\right),
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline provided $\EVV[a \sim \pol]{b(s,a)} = 0$, for any $s\in\Sspace$.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r_h^n - b\right).
\end{align*}

@@

\section{Applicability to Gaussian Policies}\label{app:gauss}
We provide more details on the applicability of Theorem \ref{theo:convergence} on the case of Gaussian policies. We start from the case of one-dimensional bounded action space $\mathcal{A}\subset\mathbb{R}$, linear mean $\mu(s) = \vtheta^T\vphi(s)$ and fixed standard deviation $\sigma$:
\[
	\pi_{\vtheta}(a\vert s) = \frac{1}{\sqrt{2\pi}\sigma}\exp\left\{
		-\frac{(\vtheta^T\vphi(s) - a)^2}{2\sigma^2}\right\},
\]
where $\vphi(s)\leq M_{\phi}$ is a bounded feature vector, and we see under which conditions the three assumptions of Section \ref{sec:conv} hold.

\boundedscore*
For the Gaussian policy defined above, it's easy to show that:
\begin{align*}
	&\nabla_{\theta_i}\log\pi_{\vtheta}(\tau) =  \phi_i(s)\frac{a-\vtheta^T\phi(s)}{\sigma^2},\\
	&\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau) = \frac{\phi_i(s)\phi_j(s)}{\sigma^2}.
\end{align*}
Hence, Assumption \ref{ass:bounded_score} is automatically satisfied \footnote{This relies on the fact that $\vtheta^T\phi(s)$ lies in bounded $\Aspace$. In practice, this is usually enforced by clipping the action selected by $\pi_{\vtheta}$. A more rigorous way would be to employ the truncated Gaussian distribution.} by taking $\GRADLOG = \frac{M_{\phi}|\Aspace|}{\sigma^2}$ and $\HESSLOG = \frac{M_{\phi}^2}{\sigma^2}$.
\par
\varreinforce*
As mentioned, \cite{pirotta2013adaptive} provides a bound on the variance of the REINFORCE estimator, adapted from \cite{zhao2011analysis}, which does not depend on $\vtheta$:
\begin{align*}
\Var\left[\gradApp{\theta_i}{N}\right] \leq \frac{R^2M_{\phi}^2H(1-\gamma^H)^2}{N\sigma^2(1-\gamma)^2}.
\end{align*}
The same authors provide a similar bound for G(PO)MDP.

\varweights*
It is noted in \cite{cortes2010learning} that, for any two Gaussian distributions $\mathcal{N}(\mu_1,\sigma_1)$ and $\mathcal{N}(\mu_2,\sigma_2)$, the variance of the importance weights from the latter to the former is bounded whenever $\sigma_2 > \frac{\sqrt{2}}{2}\sigma_1$. This is automatically satisfied by our fixed-variance Gaussian policies, since $\sigma_2=\sigma_1=\sigma$.

We now briefly examine some generalizations of the simple Gaussian policy defined above that can be found in applications:

\paragraph{Multi-dimensional actions}
When actions are multi-dimensional, factored Gaussian policies are typically employed, so the results extends trivially from the one-dimensional case. Actual multi-variate Gaussian distributions would require more calculations, but we do not expect substantially different results.

\paragraph{Non-linear mean}
In complex continuous tasks, $\mu(s)$ often represents a deep neural network, or multi-layer perceptron, where $\vtheta$ are the weights of the network. The analysis of first and second order log-derivatives in such a scenario is beyond the scope of this paper.

\paragraph{Adaptive variance}
It is a common practice to learn also the variance of the policy in order to adapt the degree of exploration. The variance (or diagonal covariance matrix in the multi-dimensional case) can be learned as a separate parameter or be state-dependent like the mean. In any case, adaptive variance must be carefully employed since it can clearly undermine all the three assumptions of Theorem \ref{theo:convergence}.
